\textbf{Urban mobility and socioeconomic deprivation in Latin America after COVID-19}

### 2024-02-12 manuscript/manuscript.tex
checking-representativeness

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{lipsum} \usepackage{libertine}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Urban mobility and socioeconomic deprivation in Latin America after COVID-19},
  pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
The movement of people between locations is crucial for sustainable and
inclusive cities, facilitating the exchange of knowledge and resources.
Due to restrictions, lockdowns and fear of crowded places, the COVID-19
pandemic significantly reduced the number and extent of people's
movements. Existing research suggests that the impact of the pandemic on
mobility was unequal across locations and socioeconomic groups. However,
evidence and analysis of long-term changes in mobility is scarce,
especially in less developed countries. In this study, we use location
data from Meta-Facebook users to analyse patterns of urban mobility in X
Functional Urban Areas from South American countries. Findings reveal a
general decrease in mobility during the pandemic, with gradual recovery
trends up to April 2023. However, socioeconomic disparities are present
through the period of analysis, with deprived areas experiencing smaller
initial losses in mobility. These disparities generally diminish over
time, although they persist in all the countries included in the
analysis . Our research highlights the importance of timely mobility
data with high spatiotemporal resolution to understand the long-term
effects of the pandemic and inform equitable policy responses that
address societal challenges in urban areas.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, breakable, sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi

\hypertarget{sec-intro}{%
\section{Introduction}\label{sec-intro}}

@@

\begin{itemize}
\tightlist
\item
  Aim - focus on Latin American countries as case study.
\item
  Aim - focus on urban areas, as the future of humanity is the future of
  cities and Latin Ameriaca is highly urbanised
\end{itemize}

\hypertarget{sec-results}{%
\section{Results}\label{sec-results}}

The evolution of the percentage change in the number of movements is
measured with respect to a baseline period prior to the pandemic as
described in Section~\ref{sec-methods}. For the purposes of the
analysis, we aggregate the raw movement data temporally into months and
spatially into administrative units at various levels according to GADM,
the Database of Global Administrative Areas\textsuperscript{32}. The
analysis focuses on administrative units that are within the boundaries
of functional urban areas as specified by the Global Human Settlement
Layer (GHSL). For each administrative unit, we compute the Relative
Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and
Applications Centre (SEDAC). Figure 1 displays the administrative units
included in the study, coloured according to their average RDI.
Predictions about the evolution of the percentage change in the number
of movements are made using the Prophet forecasting procedure. Further
details are provided in Section~\ref{sec-methods}.

\begin{figure}

@@

\end{figure}

\hypertarget{the-heterogeneous-impact-of-covid-19-on-urban-mobility}{%
\subsection{The heterogeneous impact of COVID-19 on urban
mobility}\label{the-heterogeneous-impact-of-covid-19-on-urban-mobility}}

We analyse the evolution of the percentage change in the number of
movements with respect to a baseline period prior to the pandemic.
Specifically, we focus the analysis on short-distance movements in urban
areas to represent local and routine mobility\textsuperscript{33}, so
only movements covering a distance of at most 70 km are considered. For
a movement to be classified as urban it needs to start or end within a
functional urban area from Argentina, Chile, Colombia and Mexico. The
observed data is available for a two-year period starting in April 2020,
just after the first wave of COVID-19 pandemic cases, and ending in
March 2022. After 2022 no observations are available, however, we
generate a 12-month forecast up to March 2023 in order to gain a better
understanding of the recovery trends.

@@

\hypertarget{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}{%
\subsection{The most deprived areas experienced the smallest drop in
mobility relative to pre-pandemic
levels}\label{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}}

@@

\hypertarget{sec-discussion}{%
\section{Discussion}\label{sec-discussion}}

@@

We focused the analysis on short-distance movements in urban areas,
especifically those covering 70 km or less. These journeys are typically
considered to represent local and routine mobility\textsuperscript{33} .
However, due to the characteristics of the Meta-Facebook movement data,
we are unable to distinguish the purpose of these short-distance
movements. Hence, some of our data could be capturing journeys that
involve a permanent change of place of residence. Our work therefore
motivates the need to answer questions regarding the validity of digital
footprint data for the analysis of human mobility. Further research
should focus on inferring more specific information about the nature of
the journeys, following similar approaches to those proposed
by\textsuperscript{37}, and quantifying the extent to which the digital
footprint data mirrors the true mobility patterns.

Conducting research on urban mobility using digital footprint data is
not straightforward, due to the challenges in accessing and working with
unstructured data sets which are often subject to biases and statistical
representation issues. These biases often arise from inequalities in
access and usage of digital technologies across demographic
groups\textsuperscript{38}. Despite these challenges, the data and
analysis that we used for this work provide evidence for non-trivial
patterns that are consistent across four countries in Latin America and
with other parts of the world. Our findings highlight the dynamic
interplay between socioeconomic status and urban mobility, and shall be
used to motivate and inform the public debate regarding the deep
societal consequences of urban mobility disparities on the wider
socioeconomic landscape of Latin American countries.

In conclusion, we argue that this work goes beyond the analysis of
specific patterns by demonstrating the potential of digital footprint
data for policy-relevant research on human mobility at an unprecedented
level of spatiotemporal granularity. While we have seen a rise in
initiatives to improve data services and methodological frameworks to
facilitate the use of digital footprint data for social good, progress
is still limited, especially in some parts of the world including Latin
America. It is in the hands of governments and public organisations to
prioritise the maximisation the societal benefits that digital footprint
data has to offer. This includes engaging in activities such as building
strategic partnerships with commercial data-holders and academic
institutions to establish a unified framework for the use of digital
footprint data in policy and research. In particular, we call for the
creation of resources like those developed by the European Commission
Joint Research Centre\textsuperscript{39} and the UN Statistics
Division\textsuperscript{40}, which identify sources of non-traditional
data and set methodological protocols for incorporating mobile phone
data into official mobility statistics. While current resources tend to
have a global reach, we advocate for more tailored local initiatives
that acknowledge disparities in regional data availability and adoption
of digital technologies.

\hypertarget{sec-methods}{%
\section{Methods}\label{sec-methods}}

\hypertarget{meta-facebook-movement-data}{%
\subsection{Meta-Facebook movement
data}\label{meta-facebook-movement-data}}

To capture population movements, we used anonymised aggregate mobile
phone location data from Meta users for Argentina, Colombia, Chile and
Mexico, covering a 24-month period from April 2020 to March 2022. We
used the dataset Facebook Movements created by Meta and accessed through
their Data for Good Initiative (\url{https://dataforgood.facebook.com}).
The data are built from Facebook app users who have the location
services setting turned on on their smartphone. Prior to releasing the
datasets, Meta ensures privacy and anonymity by removing personal
information and applying privacy-preserving
techniques\textsuperscript{41}. Small-count dropping is one of these
techniques. A data entry is removed if the population or movement count
for an area is lower than 10. The removal of small counts may mean that
population counts in small sparsely populated areas are not captured. A
second technique consists in adding a small undisclosed amount of random
noise to ensure that it is not possible to ascertain precise, true
counts for sparsely populated locations. Third, spatial smoothing using
inverse distance-weighted averaging is also applied is applied to
produce a smooth population count surface. The Facebook Movements
dataset offers information on the total number of Facebook users moving
between and within spatial units in the form of origin-destination
matrices. The data is temporally aggregated into three daily 8-hour time
windows (i.e.~00:00-08:00, 08:00-16:00 and 16:00-00:00). The dataset
includes a baseline capturing the number of movements before COVID-19
based on a 45-day period ending on March 10th 2020. The baseline is
computed using an average for the same time of the day and day of the
week in the period preceding March 10th. For instance, the baseline for
Monday 00:00-08:00 time window is obtained by averaging over data
collected on Mondays from 00:00 to 8:00 for the 45-day period. Details
about the baseline can be found in\textsuperscript{41}. The Bing Maps
Tile System developed by Microsoft (Microsoft) is used a spatial
framework to organise the data. The Tile System is a geospatial indexing
system that partitions the world into tile cells in a hierarchical way,
comprising 23 different levels of detail (Microsoft). At the lowest
level of detail (Level 1), the world is divided into four tiles with a
coarse spatial resolution. At each successive level, the resolution
increases by a factor of two. The data that we used are spatially
aggregated into Bing tile levels 13. That is about 4.9 x 4.9km at the
Equator\textsuperscript{41}.

\hypertarget{spatiotemporal-data-aggregation}{%
\subsection{Spatiotemporal data
aggregation}\label{spatiotemporal-data-aggregation}}

@@

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-ackers2005}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{Ackers, L.
\href{https://doi.org/10.1111/j.1468-2435.2005.00343.x}{Moving People
and Knowledge: Scientific Mobility in the European Union1}.
\emph{International Migration} \textbf{43}, 99--131 (2005).}

\leavevmode\vadjust pre{\hypertarget{ref-chen2016}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{Chen, C., Ma, J., Susilo, Y., Liu, Y. \& Wang, M.
\href{https://doi.org/10.1016/j.trc.2016.04.005}{The promises of big
data and small data for travel behavior (aka human mobility) analysis}.
\emph{Transportation Research Part C: Emerging Technologies}
\textbf{68}, 285--299 (2016).}

\leavevmode\vadjust pre{\hypertarget{ref-belik2011}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{Belik, V., Geisel, T. \& Brockmann, D.
\href{https://doi.org/10.1103/physrevx.1.011001}{Natural Human Mobility
Patterns and Spatial Spread of Infectious Diseases}. \emph{Physical
Review X} \textbf{1}, (2011).}

\leavevmode\vadjust pre{\hypertarget{ref-klugman2009human}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Klugman, J. Human development report 2009. Overcoming
barriers: Human mobility and development. \emph{Overcoming Barriers:
Human Mobility and Development (October 5, 2009). UNDP-HDRO Human
Development Reports} (2009).}

\leavevmode\vadjust pre{\hypertarget{ref-barbosa2018}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Barbosa, H. \emph{et al.}
\href{https://doi.org/10.1016/j.physrep.2018.01.001}{Human mobility:
Models and applications}. \emph{Physics Reports} \textbf{734}, 1--74
(2018).}

\leavevmode\vadjust pre{\hypertarget{ref-chinazzi2020}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Chinazzi, M. \emph{et al.}
\href{https://doi.org/10.1126/science.aba9757}{The effect of travel
restrictions on the spread of the 2019 novel coronavirus (COVID-19)
outbreak}. \emph{Science} \textbf{368}, 395--400 (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-nouvellet2021}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Nouvellet, P. \emph{et al.}
\href{https://doi.org/10.1038/s41467-021-21358-2}{Reduction in mobility
and COVID-19 transmission}. \emph{Nature Communications} \textbf{12},
(2021).}

\leavevmode\vadjust pre{\hypertarget{ref-rowe2023}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Rowe, F., González-Leonardo, M. \& Champion, T. Virtual
special issue: Internal migration in times of COVID{-}19.
\emph{Population, Space and Place} (2023)
doi:\href{https://doi.org/10.1002/psp.2652}{10.1002/psp.2652}.}

\leavevmode\vadjust pre{\hypertarget{ref-abdullah2020}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Abdullah, M., Dias, C., Muley, D. \& Shahin, Md.
\href{https://doi.org/10.1016/j.trip.2020.100255}{Exploring the impacts
of COVID-19 on travel behavior and mode preferences}.
\emph{Transportation Research Interdisciplinary Perspectives}
\textbf{8}, 100255 (2020).}

\leavevmode\vadjust pre{\hypertarget{ref-bonaccorsi2020}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Bonaccorsi, G. \emph{et al.}
\href{https://doi.org/10.1073/pnas.2007658117}{Economic and social
consequences of human mobility restrictions under COVID-19}.
\emph{Proceedings of the National Academy of Sciences} \textbf{117},
15530--15535 (2020).}

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{lipsum} \usepackage{libertine}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Urban mobility and socioeconomic deprivation in Latin America after COVID-19},
  pdfauthor={Carmen Cabrera-Arnau; Francisco Rowe; Miguel González-Leonardo; Andrea Nasuto; Ruth Neville},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\date{}
\begin{document}
\maketitle
\begin{abstract}
COVID-19 triggered a reduction in the number and extent of peoples
movements. Existing evidence suggests that while the impact of the
pandemic on mobility was widespread, the extent of this impact was
unequally felt across socioeconomic groups in the early stages of the
pandemic. Here, we find that in the long term, the most deprived
locations have experienced a more accelerated recovery towards
pre-pandemic levels of mobility. Furthermore, the socioeconomic
disparities in the patterns of mobility triggered by the first outbreak
of COVID-19, have persisted as of April 2023. These findings are based
on the analysis of time-series mobility data corresponding to X urban
areas from Latin American countries collected from Meta-Facebook users
upon their consent. Our research highlights the importance of timely
mobility data with high spatiotemporal resolution to understand the
long-term effects of the pandemic and to inform equitable policy
responses that address societal challenges in urban areas.
\end{abstract}

\hfill\break

\section{Introduction}\label{sec-intro}

@@

Latin America provides an ideal test-bed for testing these hypotheses
because of its exceptionally high levels of
inequalities\textsuperscript{32,33} and
urbanisation\textsuperscript{34}. Half of the 20 most unequal countries
in the planet are in this region. The average income Gini index of the
region is 4 percentage points higher than that of Africa and 11 higher
than China\textsuperscript{35}, and cities display some of the starkest
inequalities\textsuperscript{36}. Currently, over 80\% of the population
in Latin America live in urban areas. By 2050, this share is predicted
to reach 89\%, with the largest share concentrating in a few
megacities\textsuperscript{36}. Developing an understanding of human
mobility in Latin America is thus important to support sustainable and
inclusive spaces\textsuperscript{36}.

\section{Results}\label{sec-results}

The evolution of the percentage change in the number of movements is
measured with respect to a baseline period prior to the pandemic as
described in Section~\ref{sec-methods}. For the purposes of the
analysis, we aggregate the raw movement data temporally into months and
spatially into administrative units at various levels according to GADM,
the Database of Global Administrative Areas\textsuperscript{37}. The
analysis focuses on administrative units that are within the boundaries
of functional urban areas as specified by the Global Human Settlement
Layer (GHSL). For each administrative unit, we compute the Relative
Deprivation Index (RDI) based on data from NASA's Socioeconomic Data and
Applications Centre (SEDAC). Figure 1 displays the administrative units
included in the study, coloured according to their average RDI.
Predictions about the evolution of the percentage change in the number
of movements are made using the Prophet forecasting procedure. Further
details are provided in Section~\ref{sec-methods}.

\begin{figure}[H]

@@

\end{figure}%

\subsection{The heterogeneous impact of COVID-19 on urban
mobility}\label{the-heterogeneous-impact-of-covid-19-on-urban-mobility}

We analyse the evolution of the percentage change in the number of
movements with respect to a baseline period prior to the pandemic.
Specifically, we focus the analysis on short-distance movements in urban
areas to represent local and routine mobility\textsuperscript{38}, so
only movements covering a distance of at most 70 km are considered. For
a movement to be classified as urban it needs to start or end within a
functional urban area from Argentina, Chile, Colombia and Mexico. The
observed data is available for a two-year period starting in April 2020,
just after the first wave of COVID-19 pandemic cases, and ending in
March 2022. After 2022 no observations are available, however, we
generate a 12-month forecast up to March 2023 in order to gain a better
understanding of the recovery trends.

@@

\subsection{The most deprived areas experienced the smallest drop in
mobility relative to pre-pandemic
levels}\label{the-most-deprived-areas-experienced-the-smallest-drop-in-mobility-relative-to-pre-pandemic-levels}

@@

\section{Discussion}\label{sec-discussion}

@@

We focused the analysis on short-distance movements in urban areas,
especifically those covering 70 km or less. These journeys are typically
considered to represent local and routine mobility\textsuperscript{38} .
However, due to the characteristics of the Meta-Facebook movement data,
we are unable to distinguish the purpose of these short-distance
movements. Hence, some of our data could be capturing journeys that
involve a permanent change of place of residence. Our work therefore
motivates the need to answer questions regarding the validity of digital
footprint data for the analysis of human mobility. Further research
should focus on inferring more specific information about the nature of
the journeys, following similar approaches to those proposed
by\textsuperscript{42}, and quantifying the extent to which the digital
footprint data mirrors the true mobility patterns.

Conducting research on urban mobility using digital footprint data is
not straightforward, due to the challenges in accessing and working with
unstructured data sets which are often subject to biases and statistical
representation issues. These biases often arise from inequalities in
access and usage of digital technologies across demographic
groups\textsuperscript{43}. Despite these challenges, the data and
analysis that we used for this work provide evidence for non-trivial
patterns that are consistent across four countries in Latin America and
with other parts of the world. Our findings highlight the dynamic
interplay between socioeconomic status and urban mobility, and shall be
used to motivate and inform the public debate regarding the deep
societal consequences of urban mobility disparities on the wider
socioeconomic landscape of Latin American countries.

In conclusion, we argue that this work goes beyond the analysis of
specific patterns by demonstrating the potential of digital footprint
data for policy-relevant research on human mobility at an unprecedented
level of spatiotemporal granularity. While we have seen a rise in
initiatives to improve data services and methodological frameworks to
facilitate the use of digital footprint data for social good, progress
is still limited, especially in some parts of the world including Latin
America. It is in the hands of governments and public organisations to
prioritise the maximisation the societal benefits that digital footprint
data has to offer. This includes engaging in activities such as building
strategic partnerships with commercial data-holders and academic
institutions to establish a unified framework for the use of digital
footprint data in policy and research. In particular, we call for the
creation of resources like those developed by the European Commission
Joint Research Centre\textsuperscript{44} and the UN Statistics
Division\textsuperscript{45}, which identify sources of non-traditional
data and set methodological protocols for incorporating mobile phone
data into official mobility statistics. While current resources tend to
have a global reach, we advocate for more tailored local initiatives
that acknowledge disparities in regional data availability and adoption
of digital technologies.

\section{Methods}\label{sec-methods}

\subsection{Meta-Facebook movement
data}\label{meta-facebook-movement-data}

To capture population movements, we used anonymised aggregate mobile
phone location data from Meta users for Argentina, Colombia, Chile and
Mexico, covering a 24-month period from April 2020 to March 2022. We
used the dataset Facebook Movements created by Meta and accessed through
their Data for Good Initiative (\url{https://dataforgood.facebook.com}).
The data are built from Facebook app users who have the location
services setting turned on on their smartphone. Prior to releasing the
datasets, Meta ensures privacy and anonymity by removing personal
information and applying privacy-preserving
techniques\textsuperscript{46}. Small-count dropping is one of these
techniques. A data entry is removed if the population or movement count
for an area is lower than 10. The removal of small counts may mean that
population counts in small sparsely populated areas are not captured. A
second technique consists in adding a small undisclosed amount of random
noise to ensure that it is not possible to ascertain precise, true
counts for sparsely populated locations. Third, spatial smoothing using
inverse distance-weighted averaging is also applied is applied to
produce a smooth population count surface. The Facebook Movements
dataset offers information on the total number of Facebook users moving
between and within spatial units in the form of origin-destination
matrices. The data is temporally aggregated into three daily 8-hour time
windows (i.e.~00:00-08:00, 08:00-16:00 and 16:00-00:00). The dataset
includes a baseline capturing the number of movements before COVID-19
based on a 45-day period ending on March 10th 2020. The baseline is
computed using an average for the same time of the day and day of the
week in the period preceding March 10th. For instance, the baseline for
Monday 00:00-08:00 time window is obtained by averaging over data
collected on Mondays from 00:00 to 8:00 for the 45-day period. Details
about the baseline can be found in\textsuperscript{46}. The Bing Maps
Tile System developed by Microsoft (Microsoft) is used a spatial
framework to organise the data. The Tile System is a geospatial indexing
system that partitions the world into tile cells in a hierarchical way,
comprising 23 different levels of detail (Microsoft). At the lowest
level of detail (Level 1), the world is divided into four tiles with a
coarse spatial resolution. At each successive level, the resolution
increases by a factor of two. The data that we used are spatially
aggregated into Bing tile levels 13. That is about 4.9 x 4.9km at the
Equator\textsuperscript{46}.

\subsection{Spatiotemporal data
aggregation}\label{spatiotemporal-data-aggregation}

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{0}{0}
\bibitem[\citeproctext]{ref-ackers2005}
\CSLLeftMargin{1. }%
\CSLRightInline{Ackers, L.
\href{https://doi.org/10.1111/j.1468-2435.2005.00343.x}{Moving People
and Knowledge: Scientific Mobility in the European Union1}.
\emph{International Migration} \textbf{43}, 99--131 (2005).}

\bibitem[\citeproctext]{ref-chen2016}
\CSLLeftMargin{2. }%
\CSLRightInline{Chen, C., Ma, J., Susilo, Y., Liu, Y. \& Wang, M.
\href{https://doi.org/10.1016/j.trc.2016.04.005}{The promises of big
data and small data for travel behavior (aka human mobility) analysis}.
\emph{Transportation Research Part C: Emerging Technologies}
\textbf{68}, 285--299 (2016).}

\bibitem[\citeproctext]{ref-belik2011}
\CSLLeftMargin{3. }%
\CSLRightInline{Belik, V., Geisel, T. \& Brockmann, D.
\href{https://doi.org/10.1103/physrevx.1.011001}{Natural Human Mobility
Patterns and Spatial Spread of Infectious Diseases}. \emph{Physical
Review X} \textbf{1}, (2011).}

\bibitem[\citeproctext]{ref-klugman2009human}
\CSLLeftMargin{4. }%
\CSLRightInline{Klugman, J. Human development report 2009. Overcoming
barriers: Human mobility and development. \emph{Overcoming Barriers:
Human Mobility and Development (October 5, 2009). UNDP-HDRO Human
Development Reports} (2009).}

\bibitem[\citeproctext]{ref-barbosa2018}
\CSLLeftMargin{5. }%
\CSLRightInline{Barbosa, H. \emph{et al.}
\href{https://doi.org/10.1016/j.physrep.2018.01.001}{Human mobility:
Models and applications}. \emph{Physics Reports} \textbf{734}, 1--74
(2018).}

\bibitem[\citeproctext]{ref-chinazzi2020}
\CSLLeftMargin{6. }%
\CSLRightInline{Chinazzi, M. \emph{et al.}
\href{https://doi.org/10.1126/science.aba9757}{The effect of travel
restrictions on the spread of the 2019 novel coronavirus (COVID-19)
outbreak}. \emph{Science} \textbf{368}, 395--400 (2020).}

\bibitem[\citeproctext]{ref-nouvellet2021}
\CSLLeftMargin{7. }%
\CSLRightInline{Nouvellet, P. \emph{et al.}
\href{https://doi.org/10.1038/s41467-021-21358-2}{Reduction in mobility
and COVID-19 transmission}. \emph{Nature Communications} \textbf{12},
(2021).}

\bibitem[\citeproctext]{ref-rowe2023}
\CSLLeftMargin{8. }%
\CSLRightInline{Rowe, F., González-Leonardo, M. \& Champion, T. Virtual
special issue: Internal migration in times of COVID{-}19.
\emph{Population, Space and Place} (2023)
doi:\href{https://doi.org/10.1002/psp.2652}{10.1002/psp.2652}.}

\bibitem[\citeproctext]{ref-abdullah2020}
\CSLLeftMargin{9. }%
\CSLRightInline{Abdullah, M., Dias, C., Muley, D. \& Shahin, Md.
\href{https://doi.org/10.1016/j.trip.2020.100255}{Exploring the impacts
of COVID-19 on travel behavior and mode preferences}.
\emph{Transportation Research Interdisciplinary Perspectives}
\textbf{8}, 100255 (2020).}

\bibitem[\citeproctext]{ref-bonaccorsi2020}
\CSLLeftMargin{10. }%
\CSLRightInline{Bonaccorsi, G. \emph{et al.}
\href{https://doi.org/10.1073/pnas.2007658117}{Economic and social
consequences of human mobility restrictions under COVID-19}.
\emph{Proceedings of the National Academy of Sciences} \textbf{117},
15530--15535 (2020).}

\bibitem[\citeproctext]{ref-abu-rayash2020}
\CSLLeftMargin{11. }%
\CSLRightInline{Abu-Rayash, A. \& Dincer, I.
\href{https://doi.org/10.1016/j.erss.2020.101693}{Analysis of mobility
trends during the COVID-19 coronavirus pandemic: Exploring the impacts
on global aviation and travel in selected cities}. \emph{Energy Research
\& Social Science} \textbf{68}, 101693 (2020).}

\bibitem[\citeproctext]{ref-lee2023}
\CSLLeftMargin{12. }%
\CSLRightInline{Lee, S., Ko, E., Jang, K. \& Kim, S.
\href{https://doi.org/10.1016/j.cities.2023.104223}{Understanding
individual-level travel behavior changes due to COVID-19: Trip
frequency, trip regularity, and trip distance}. \emph{Cities}
\textbf{135}, 104223 (2023).}
